---
title: "Data structures and data cleaning"
author: "Yuan Li"
email: "yuanliutk@gmail.com"
date: "2023-09-27"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{=html}
<style type="text/css">
body {
font-size: 12pt;
}

</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Data types

The main data type in R is vector, list, matrix and table. For each type, how to create, slice and what functions to apply are very important.

### 1.1 String

#### 1.1.1 Basic functions of string

Understanding string needs the knowledge of vector and list.

```{r}
library(stringr)


# Remove the white space
str_1 <- " Hi WORLD "
typeof(str_1)
str_trim(str_1)
str_1 %>% str_trim

# Change upper and lower case
str_1 %>% str_trim %>% str_to_lower
str_1 %>% str_trim %>% str_to_upper
str_1 %>% str_trim %>% str_to_title


# Find element
str_1 %>% str_detect(" ")

# Replace 
str_1 %>% str_replace(" ", '')

"I want to lrn R, and I definitely wnt to lrn more." %>%
  str_replace_all('lrn', 'learn')

# Use regex to replace
str_3  <-  "aBcDEfGHIj"
str_replace(str_3, '[A-Z]', '-')
str_replace_all(str_3, '[A-Z]', '-')

sql_variations  <- c("SQL", "Sql", "sql")
sql_uniform  <-  str_replace(sql_variations, "(?i)sql", "SQL")
sql_uniform


email_variations  <-  c('email', 'Email', 'e Mail',
                        'e mail', 'E-mail', 'e-mail',
                        'eMail', 'E-Mail', 'EMAIL')

pattern <- "(?i)e[\\s\\-]?mail"
email_uniform <- str_replace_all(email_variations, pattern, 'email') 
email_uniform

# Padding
"data" %>% str_pad(width=20, side="both", pad=" ")

# Split
str_2 <- "I want to lrn R, and I definitely wnt to lrn more." %>%
  str_split(' ')
str_2
typeof(str_2)

str_2[[1]][1:3]


# Extract string values

str_4 <- 'nlass123'
str_sub(str_4, -3, -1)
str_sub(str_4, 6, -1)

# Match

str_5 <- "Python 3.7"
str_match(str_5, "\\d.\\d")[1]


# Covert form vector to string

words <- c("I", "Have", "A", "Dream")
words

str_c(words, collapse=" ")

```

After split, the data type change from character to list. To slice it, needs to use [[]] to get the value and then slice it. The details can be seen in Section 1.3 `List`.

#### 1.1.2 Time series

```{r}
library(lubridate)
date1 <- "01011997"
date2 <- "2019-03-04"


date_1_s <- mdy(date1)
date_1_s

date_2_s <- ymd(date2)
date_2_s


year(date_2_s)
month(date_2_s)
day(date_2_s)
week(date_2_s)

date_1 <- "2020-04-01 14:20:59 EDT"
date_1 %>%  ymd_hms(tz='EDT') 

date_1 %>% wday()
date_1 %>% yday()


# Duration and period
dur <- duration(day=1, hour=1, minute=30)
perd <- period(years=1, months=1, days=1)
ymd("2020-04-01") + perd
ymd("2020-04-01") + dur

```

### 1.2 Vectors

### 1.2.1 Define the vector

```{r}
# Use colons
range_a <- 1:10
range_a

```

```{r}
# Use function seq
seq_a <- seq(1, 10, 2)
seq_a

# repeated values
rep(1:3, 2)

```

```{r}
# Use c function
c_a <- c(1, 2, 3, 4)
c_a
print(paste("Type of c_a: ", typeof(c_a)))

# Type of 1L is integer.
c_a_int <- c(1L, 2L, 3L, 4L)
c_a_int
print(paste("Type of c_a_int: ", typeof(c_a_int)))

```

```{r}
# Logical vector
is_a <- c(TRUE, FALSE, TRUE, FALSE)
is_a 
print(paste("Type of is_a is: ", typeof(is_a)))

# Use logical vector to select values:
c_a[is_a]

```

### 1.2.2 Operations of vector

#### 1.2.2.1 Scaling up or down

-   Appending

```{r}
# Appending
combine <- c(range_a, c_a, seq_a, is_a )
combine
typeof(combine)

```

When logical vector is combined, TRUE is converted into 1 and FALSE to 0.

-   Slicing and Removing

```{r}
 
# slicing: make the interested indices into a vector
combine[range_a]
combine[seq_a]
combine[is_a]

combine[c(TRUE)]
combine[c(TRUE, FALSE)]
combine[c(TRUE, FALSE, TRUE)]
combine[c(TRUE, FALSE, TRUE, FALSE)]

# Removing: rule out the indices not interested, use a negative symbol
combine[-range_a]
combine[-seq_a]
combine[-is_a]


```

For numeric vector, the selected indices are the values in the vector. For logical vector, it is a state, one or two value is enough. If the vector is TRUE, it means all the indices are selected. [TRUE, FALSE], odd indices are selected. [TRUE, FALSE, TRUE, FALSE] has the same slicing as [TRUE, FALSE]

#### 1.2.2.2 Mathematical functions

```{r}
# Length
library(dplyr)
length(combine)
combine %>% length
print(paste("The length of vector combine is: ", length(combine)))

# Unique values
unique(combine)
combine %>% unique

# Count of each element
table_c <- combine %>% table 
table_c 
table_c[1]


# Sum of all the elements
sum(combine)
print(paste("The sum of the elements in vector combine are: ", sum(combine)))


# Mean of all the elements
mean(combine)
print(paste("The mean of the elements in vector combine are: ", mean(combine)))

# Sort
sort(range_a)


# Use statement to filter
range_a[range_a>3] # In essence using logical vector

```

#### 1.2.2.3 Giving a name to each eleent

```{r}

names(c_a) <- c("A", "B", 'C', 'D')
typeof(c_a)

# Using the indices to slice or removing
c_a[c(1,3)]
c_a[-c(1,3)]

# Appending a vector, the vector won't have a name
c(c_a, E=10)

# Get the value without name
c_a[[1]]
c_a[['A']]



```

To obtain the vector value, it can only extract one value. Because the type after naming is the same, the functions still work.

### 1.3 List

In R, an additional attribute of the element in a vector can be assigned. It is named list. Using different slicing strategy, the list or the vector can be sliced.

```{r}
# Creating a list, or appending
list(seq_a, range_a)

list_a <- list(seq = seq_a, range = range_a)
list_a
typeof(list_a)


# Slicing a list
list_a[1]  
list_a[[1]]  

list_a['seq']
typeof(list_a['seq'])

list_a[['seq']]
typeof(list_a[['seq']]) 

# Removing, cannot use the nage

list_c <- list(seq = seq_a, range = range_a, kit=c(1,2,58))
list_c



list_c[-c(1)]

# Get the vector 

list_c[-c(1)][[1]]

# Get the cell
list_c[-c(1)][[c(2, 3)]]

list_c[TRUE]
list_c[c(TRUE, FALSE)]
list_c

# Unlist
list_c[1]
unlist(list_c[1])

```

-   The name can be used to get the specific column: list and vector.
-   The column numbers can be used to slicing and removing one/multiple columns. But the vector can only be extracted single column or one cell.
-   Using [] for list, [[]] for vector.
-   Only the vector can use the mathematical functions.

### 1.4 Matrix

#### 1.4.1 Create matrix

-   Using matrix function to create matrix

```{r}
# Create matrix
mat <- matrix(1:2, nrow = 2, ncol = 3)
mat

typeof(mat)

# Extract the value
mat[1, ]
mat[,1]


# Names

colnames(mat) <- c('A', 'B', 'C')
mat

# 
typeof(mat)

mat[1, 'B']


```

-   Using cbind or rbind function to create matrix

```{r}

harvard <- c(1,1,1,1,3)
stanford <- c(2,9,3,4,10)
MIT <- c(3,3,2,2,1)
cambridge <- c(4,2,6,13,48)
oxford <- c(5,7,12,9,15)
columbia <- c(6,13,13,12,4)
university_matrix <- rbind(harvard, stanford, MIT, cambridge, oxford, columbia)
university_matrix

# The vector name shown in the row name
typeof(university_matrix)
colnames(university_matrix) <- c("world_rank", "quality_of_education", "influence", "broad_impact", "patents")

university_matrix
rownames(university_matrix)



```

-   The type of matrix is double or integer, meaning it is numeric.
-   cbind or rbind are used to bind vectors. It can be used to append a vector to an existing matrix. bind_rows, bind_cols are to bind the rows or cols in a dataframe

#### 1.4.2 Slice matrix

Because matrix is 2-d array. We need to assign both the row and col to get locate the elements. Like slicing vector and list, we can use a vector of number, boolean, or rownames/colnames.

```{r}
# Vector of rownames, colnames
university_matrix[c("harvard", "stanford", "MIT"), c("world_rank", "quality_of_education")]
university_matrix[c("harvard", "stanford", "MIT"), ]


# Vector of indices
university_matrix[1: 3, 1:2]
university_matrix[, 1]
university_matrix[1:3, ]

# Vector of indices
university_matrix[c(TRUE, FALSE), c(TRUE, FALSE)]
university_matrix[, c(TRUE, FALSE)]

# Mixed
university_matrix[c("harvard", "stanford", "MIT"), 1:2]


# Conditional statement
university_matrix[university_matrix[, 'patents'] > 5, ]


# Remove
university_matrix[ -c(1, 2), ]
university_matrix[ -c(1, 2), -c(2,3)]

university_matrix[ c(FALSE, FALSE, TRUE), c(TRUE, FALSE, TRUE)]

```

To remove the slices, numeric vector is allowed. Or set the row or column to remove to be FALSE.

#### 1.4.3 Mathmatical operation

```{r}
# Whole matrix
sum(university_matrix)
mean(university_matrix)

# Rows 
rowSums(university_matrix)
rowMeans(university_matrix)

# Columns
colSums(university_matrix)
colMeans(university_matrix)

# Rank
rank(university_matrix)

# Summary
summary(university_matrix)


```

### 1.5 Dataframe

#### 1.5.1 Create Dataframe

The result from read csv and excel is a dataframe. But also, using other functions can also create data frame. So many functions are available to clean dataframe.

```{r}

# data.frame
data.frame(x=1:5, y=2:6)

# as.tibble
library(tibble)
df_tibble <- as.tibble(university_matrix, rownames=' ')
df_tibble


# As.data.frame
df <- as.data.frame(university_matrix)
df
typeof(df)


# Append 
tt <- list(name=c(1,2,4, 8, 10, 11))
bind_cols(df, tt)

pp <- tibble(name=c(1,2,4, 8, 10, 11))
bind_cols(df, pp)

```

#### 1.5.2 Slice data

```{r}

# Obtain small amount of data
head(df, 10) # default is 5 
tail(df, 10) # default is 5


library(dplyr)
sample(df, 10, replace=TRUE)
sample_frac(df, 0.1)

# Select the columns

df$world_rank
df[1]
df[c('world_rank', 'quality_of_education')] 

df[1:4, ]

df %>% select(world_rank , patents)
df %>% pull(world_rank, patents)

# Unselect some rows or cols.
df %>% select(1:4)
df %>% select(-1, -4)
df %>% select(-world_rank , -patents)

df %>% select(where(is.numeric))


df %>% select(world_rank , patents) %>% sum
df %>% pull(world_rank, patents) %>% sum

# Conditional statement to filter
df[df$world_rank>3, ][c(1,3)]

```

-   Type of dataframe is list. So use [[]] to obtain the values. But for two column values, pull function works.
-   Select function can select different columns.
-   Pipeline can make the process more efficiently.

#### 1.5.3 Basic functions

```{r}

# Check the dimension
dimension <- dim(df)
dimension

# Check the datatype of each column
glimpse(df)

# Check if there are missing values
colSums(is.na(df))

# Create a new column
df %>% mutate(tution=c(5000, 5500, 7800, 7900, 10000, 8100))

# Summarize to get the statistic values
df %>% summarize(sum = sum(world_rank), 
                 n = n(),
                 mean = mean(world_rank))

# Get the statistics for each column
library(tidyr)

df %>% pivot_longer(cols = c(colnames(df)), names_to = "Column", values_to = "value") %>% group_by(Column) %>% 
  summarize(sum = sum(value),
            mean=mean(value),
            median = median(value)) %>%
  arrange(-mean)

# filter
df %>% filter(patents >10)
df %>% filter(patents %in% c(1, 5, 10, 20))
df %>% filter(!patents %in% c(1, 5, 10, 20))

# Categorize the group

df %>% mutate(Class=case_when(
  world_rank <= 2 ~ 'First Class',
  world_rank <= 4 ~ 'Second Class',
  TRUE ~ 'Third Class'))

df %>% mutate(Is_first_class=if_else(world_rank <= 2,'First Class', 'Second Class'))

# Unique

unique(df$world_rank)


# duplicated

sum(duplicated(df$world_rank))

# distinct: remove all the duplicated values

df %>% distinct(world_rank, .keep_all = TRUE)

# cor: Pearson's r

df %>% select(world_rank, influence) %>% cor(use="pairwise.complete.obs")

```

#### 1.5.4 Dataframe join

```{r}

f <- list(A=c(1,3,5,6), B=c(55, 88,45,12))
g <- list(A=c(1,3,5,8, 9), C=c('Hi', 'Kitty', 'Yes',12, 'Good'))

df1 <- as.tibble(f)
df2 <- as.tibble(g)

df1 %>% inner_join(df2, by='A')
df1 %>% left_join(df2, by='A')
df1 %>% right_join(df2, by='A')


```

### 1.6 JSON

```{r}
library(jsonlite)

world_cup_str  <-  '
[
    {
        "team_1": "France",
        "team_2": "Croatia",
        "game_type": "Final",
        "score" : [4, 2]
    },
    {
        "team_1": "Belgium",
        "team_2": "England",
        "game_type": "3rd/4th Playoff",
        "score" : [2, 0]
    }
    ]
'
typeof(world_cup_str)

# fromJSON turn the json data into dataframe
world_cup_df <- fromJSON(world_cup_str)
typeof(world_cup_df)


```

## 2. Data Cleaning

To select the data we are interested in, customized functions or advanced functions has to be used. In addition, we also have to imputate the missing values, values inconsitence in our data using different functions.

### 2.1 User defined functions

```{r}
# For loop
for (i in 1:10) {
  print(i+5)
}


for (i in list_c){
  print(i)
  print(typeof(i))
}


# If-else
i =5
if (i==1) {
  print("The input is 1")
} else {
print("The input is not 1")
}


# Define a function

add_x_y <- function () { 
  t=0
  for (x in 1:10) {
    if (x%%2 == 0) {
      t = t + x
    }
    
  }
  return(t)
}

add_x_y()





# Input is a list

new_add <- function(x,y, ...) {
  return(list(...))
}
new_add(1,2, c(3, 4))


new_add <- function(x,y, ...) {
  current_sum = x + y
  for (num in list(...)) {
    current_sum = current_sum + num
  }
  return(current_sum)
}

new_add(1, 2, 3, 4)

```

### 2.2 Functions for data cleaning

#### 2.2.1 Data inconsistency

For one column, most of data is double but some are strings. It needs to make it consistent.

```{r}

# Change one column data to numeric
library(readr)
data <- tibble(test=c(1, 2, "5", 'Fail', 'None'))
data

data %>% mutate(test=as.numeric(test))
data %>% mutate(test=parse_number(test))

# Change multiple columns to be numeric
library(dplyr)


df %>% mutate(across(c('world_rank', 'influence'), as.numeric))
df %>% mutate(across(c(1, 3), as.numeric))
df %>% mutate(across(1:3, as.numeric))
df %>% mutate(across(starts_with('broad'), as.numeric))
df %>% mutate(across(contains('broad'), as.numeric))


# str replace 
email_variations  <-  c('email', 'Email', 'e Mail',
                        'e mail', 'E-mail', 'e-mail',
                        'eMail', 'E-Mail', 'EMAIL')

pattern <- "(?i)e[\\s\\-]?mail"

email_uniform <- str_replace_all(email_variations, pattern, 'email')
email_uniform
      


```

#### 2.2.2 Dealing with columns with missing values

```{r}

eg <- list(col1= c(1, 5, 7, 8, 'None'), col2=c(8, 40, 50, '5', '7'))
df5 <- as.data.frame(eg)  %>% mutate(across(1:2, parse_number))
df5

# Check the number of missing value
colSums(is.na(df5))

# Mean
colMeans(df5, na.rm=TRUE)

# Just use the non-null data
df5 %>% filter(!is.na(col1))

# drop_na
df5 %>% drop_na()

# fill na with the column average value: if_else or case_when

mean_1 <- mean(df5$col1, na.rm=TRUE)
mean_1

df5 %>% mutate(col1=if_else(is.na(col1), mean_1, col1))
df5 %>% mutate(col1=case_when(
  is.na(col1) ~ mean_1,
  TRUE ~ col1)
  )


```

#### 2.2.3 Element-wise operation: map function

1). `map` function can process the whole columns as a vector element-wise, much faster than `for` loop. The output of map function is the result of applying the function to each element in columns. The output of map is a list. To get the value, we have to use [[]]. But there are some specific function to get the value based on the data type: map_int, map_dbl, map_chr, map_lgl and map_df 2). pmap can operate on elements in several columns

```{r}
library(purrr)

new_df <- df5 %>% mutate(col1=case_when(
  is.na(col1) ~ mean_1,
  TRUE ~ col1))

new_df

# map function

map(new_df, length)

map_int(new_df, length)
map_chr(new_df, class)

# map with user-defined function

map_df(new_df, function(x) x+1) 


# pmap

data1 <- c(1, 3, 5)
data2 <- c(2, 4, 6)
data3 <- c(8, 9, 7)

data_list <- list(x=data1, y=data2, z=data3) %>% as_tibble
data_list

plus <- function(x, y, z) x + y + z # x,y,z here corresponds to the colnames

sums <- pmap_int(data_list, plus)
sums

```

## Common library and functions
1. readr:
  * read_csv   
  * read_excel   
  * parse_number
2. dplyr
  * %>% 
  * glimpse
  * filter    
  * arrange   
  * select    
  * mutate    
  * summarize   
  * group_by
3. tidyr
  * tibble    
  * pivot_longer    
  * pivot_wider
  * replace_na    
  * nest
4. stringr
  * str_sub   
  * str_match   
  * str_replace   
  * str_detect
5. purrr
  * map   
  * pmap
6. caret
  * createDataPartition   
  * train
7. broom
  * augment   
  * glance    
  * tidy
8. lubridate
  * ymd   
  * year 
  * wday   
  * period    
  * duration
9. ggplot2
  * ggplot
  
  
  

