---
title: "New York Property Price"
author: "Alex"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message=FALSE)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The data is a subset of condominium sales data from all five boroughs of New York City. 

The purpose of this Guided Project is to explore the following questions:

1. How well does the size of a condominium (measured in gross square feet) explain or predict sale price across New York City as a whole?
* To answer this, we'll explore property sales data for all five boroughs combined.
2. How well does the size of a condominium explain or predict sale price for each individual borough?
*For this question, we'll build, analyze, and compare linear models for each borough.

Along the way, we'll also see how erroneous sale price entries influence our models, and we will devise methods to remove problematic entries.


## 1. Read data
```{r}
library(readr)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(caret)
library(broom)

# Read data
ny <- read_csv('NYC_property_sales.csv',  col_types = cols())
head(ny)

# Data type
map_df(ny, typeof)


# Missing data
colSums(is.na(ny))


# Using plot, price vs gross square feet

NYC_condos <- ny %>%
  filter(building_class_at_time_of_sale=='R4')

ggplot(data=NYC_condos, aes(x=gross_square_feet/1e3, y=sale_price/1e6)) + geom_point(alpha=0.5) + 
  scale_y_continuous(labels=scales::comma)  + 
  geom_smooth(method='lm', se=FALSE) + 
  theme_bw() + 
  labs(title='Condos Sale Price vs Gross Square Feet', x='Gross Square Feet (k sqft)', y= 'Sale Price (M $)')


# Zoom-in
ggplot(data=NYC_condos, aes(x=gross_square_feet/1e3, y=sale_price/1e6)) + geom_point(alpha=0.5) + 
  scale_y_continuous(labels=scales::comma) + 
  xlim(0, 5) + 
  ylim(0, 20) + 
  geom_smooth(method='lm', se=FALSE) + 
  theme_bw() + 
  labs(title='Zoom-in Condos Sale Price vs Gross Square Feet', x='Gross Square Feet (k sqft)', y= 'Sale Price (M $)')

# Facet wrap

ggplot(data=NYC_condos, aes(x=gross_square_feet/1e3, y=sale_price/1e6)) + geom_point(alpha=0.5) + 
  scale_y_continuous(labels=scales::comma) +  
  geom_smooth(method='lm', se=FALSE) + 
  theme_bw() + 
  labs(title='Sale Price vs Gross Square Feet', x='Gross Square Feet (k sqft)', y= 'Sale Price (M $)') + 
  facet_wrap(vars(borough), ncol=2, scales='free')


```

Generally, the sale price increasing with the size. Queens seems weaker linear relationship. The outlier occurs in Manhattan, which has 25 and 200 k-sqft with price of 250 M dollars. Most of the house size are below 2000 sqft.Brooklyn has several houses with sale price 30 million. 

## 2. Data Cleaning

### 2.1 Check the details of the outlier and decide keep or drop it
```{r}
# check the price outliers

NYC_condos %>% 
  filter(sale_price >100e6) %>%
  select(address, zip_code, gross_square_feet, sale_price, sale_date)

```
1. [240 M property is located in 220 Central Park South](https://www.6sqft.com/billionaire-ken-griffin-buys-238m-nyc-penthouse-the-most-expensive-home-sold-in-the-u-s/]. It is skyscraper. It is composed of two sections: a 70-story, 950-foot (290 m) tower on 58th Street, which is the 21st-tallest building in New York City, as well as an 18-story section on Central Park South, both of which contain a limestone facade. Most of the 118 apartments are duplex apartments, although some of the units have been combined to create larger units. The building has a porte-cochère, a wine cellar, private dining rooms, and various recreational facilities
2. [200 M is the located in 165 East 66th St, Resi](https://therealdeal.com/new-york/2019/04/12/cim-group-acquires-resi-portion-of-ues-luxury-rental-for-200m/). It is a UES luxury rental tower.
These two are not commercial properties and should be cleaned.

### 2.2 Duplicated addresses houses

```{r}

dup_address <- NYC_condos %>% 
  filter(duplicated(NYC_condos$address)) %>%
  select(address, sale_price)

```
There are 40 units at 554 4 Avenue, with the same sale_price of 29620207$. It is more likely that the sale price is for 40 unit. The average gross square feet is 843 and sale price per unit is 740505 dollar. 


### 2.3 Clean these duplicted address houses
```{r}

multi_unit_sales <- NYC_condos %>% 
  group_by(sale_price, sale_date) %>%
  filter(n()>3) %>%
  arrange(sale_price) 

condos_cc <- NYC_condos %>%
  filter(sale_price < 10e6) %>%
  anti_join(multi_unit_sales)


ggplot(data=condos_cc, aes(x=gross_square_feet/1e3, y=sale_price/1e6)) + geom_point(alpha=0.5) + 
  scale_y_continuous(labels=scales::comma) +  
  geom_smooth(method='lm', se=FALSE) + 
  theme_bw() + 
  labs(title='Sale Price vs Gross Square Feet', x='Gross Square Feet (k sqft)', y= 'Sale Price (M $)') + 
  facet_wrap(vars(borough), ncol=2, scales='free')

```
After cleaning the multiple unit sales, the linear relationship is very prominent.

## 3. Linear model fitting
```{r}

condos_lm <- lm(sale_price ~ gross_square_feet, data=condos_cc)
condos_lm_org <- lm(sale_price ~ gross_square_feet, data=NYC_condos)

library(tibble)
lm_fit <- summary(condos_lm)
lm_org_fit <- summary(condos_lm_org)
lm_org_fit[2]
lm_org_fit

lm_org_fit$coefficients [, 1]
lm_org_fit$coefficients


# confident interval
confint(condos_lm)
confint(condos_lm_org)

# RSE
c(sigma(condos_lm_org), sigma(condos_lm))

lm_fit$r.square
lm_org_fit$r.square

lm_fit$adj.r.square
lm_org_fit$adj.r.square

```
Both the original and cleaned data fit reject the null hypothesis.
After cleaning, the R square increased from 0.32 to 0.66. 


## Nest functions:create multiple linear models
```{r}
# Create and nest linear model, tidy_coeff, glance_coeff, augment_coff

condos_nest <-
  condos_cc %>%
  group_by(borough) %>%
  nest() %>%
  mutate(linear_model = map(.x=data, .f= ~lm(sale_price ~ gross_square_feet, data=.))) %>%
  mutate(tidy_coef = map(.x=linear_model,.f=tidy, conf.int = TRUE )) %>%
  mutate(glance_coef=map(.x=linear_model, .f=glance)) %>%
  mutate(augment_coef=map2(.x=linear_model, .y=data, .f=augment))


# Unnest tidy_coef, tidy_slope, glance_coef  
tidy_coef <- condos_nest %>%
  select(borough, tidy_coef) %>%
  unnest(cols = tidy_coef)
tidy_coef


tidy_slope <- tidy_coef %>%
  filter(term=='gross_square_feet') %>%
  arrange(desc(estimate))

glance_coef <- condos_nest %>%
  select(borough, glance_coef) %>%
  unnest(cols=glance_coef) %>%
  arrange(desc(r.squared))
glance_coef


augment_coef <- condos_nest %>%
  select(borough, augment_coef) %>%
  unnest(cols=augment_coef)
augment_coef 

# Correlation coefficients
condos_cor <-
  condos_cc %>%
  group_by(borough) %>%
 summarize(cor= cor(gross_square_feet, sale_price)) %>%
  arrange(desc(cor))
condos_cor

# Compare correlation coefficients and RMSE

bind_cols(condos_cor, RSME=glance_coef$r.squared) %>% arrange(-cor)

```
The correlation coefficients have the same trends as RSME.